Exploring probability distributions for bivariate temporal granularities

### 2020-02-07 main.Rmd
commits to include all latex packages for including jcgs format

   Visualizing data across granularities which are either single-order-up or multiple-order-up or periodic/aperiodic helps us to understand periodicities, pattern and anomalies in the data. Because of the large volume of data available, using displays of probability distributions conditional on one or more granularities is a potentially useful approach. This work provides tools for creating granularities and exploring the associated time series within the tidy workflow, so that probability distributions can be examined using the range of graphics available in ggplot2 [@Wickham2009-pk]. 
bibliography: [bibliography.bib]
preamble: >
  \usepackage{mathtools}
  \usepackage{amssymb} 
  \usepackage{Booktabs}
output:
  bookdown::pdf_book:
    base_format: rticles::asa_article
    fig_height: 5
    fig_width: 8
    fig_caption: yes
    dev: "pdf"
---

@@

Calendar-based graphics [@wang2018calendar] are useful in visualizing patterns in the weekly and monthly structure well and are capable of checking the weekends or special days. Any sub-daily resolution temporal data can also be displayed using this type of faceting [@Wickham2009-pk] with days of the week, month of the year and another sub-daily deconstruction of time. But calendar effects are not restricted to conventional day-of-week, month-of-year ways of deconstructing time. There can be several classes of time deconstructions, viz. based on the arrangement (linear vs. cyclic) or hierarchical order of the calendar. Linear time granularities respect the linear progression of time such as hours, days, weeks and months. One of the first attempts of characterizing these granularities occur in @Bettini1998-ed. The definitions and rules defined are inadequate to reflect periodicities in time. Hence, there is a need to define cyclic time granularities in a different approach, which can be useful in visualizing periodic behavior. Cyclic time granularities can be circular, quasi-circular or aperiodic. Examples of circular can be hour of the day, day of the week, that of quasi-circular can be day of the month and examples of aperiodic granularities can be public or school holidays. Time deconstructions can also be based on the hierarchical structure of time. For example, hours are nested within days, days within weeks, weeks within months, and so on. Hence, it is possible to construct single-order-up granularities like second of the minute or multiple-order-up granularities like second of the hour. lubridate [@G_Grolemund2011-vm] creates easy access and manipulation of common date-time objects. But most accessor functions are limited to single-order-up granularities.

@@

The grammar of graphics introduced a framework to construct statistical graphics by relating data space to the graphic space [@Wilkinson1999-nk]. The layered grammar of graphics proposed by [@Wickham2009-pk], which is an alternate and modified parametrization of the grammar suggests that graphics are made up of distinct layers of grammatical elements. There are seven grammatical elements in total, out of which data, aesthetics and geometries are essential elements. Data are observations made on a number of variables, typically thought of as a sample from a population distribution. Aesthetic mapping describes how variables are mapped to elements of the plot, such as horizontal axis or color. Geometries determine the physical representation of the data in the plot. The remaining grammatical elements of a graphic namely facets, statistics, coordinates and themes are optional and contains detail of the plot. Facets dictates how to split up our plot and is a powerful tool for exploratory data analysis as we can rapidly compare patterns in different parts of the data and can see whether they are the same or different.

@@

<!-- We exploit some features of layers and faceting [@Wilkinson1999-nk] and [@Wickham2009-pk] to come up with a recommendation system while visualizing the distribution of the "dependent" variable across bivariate cyclic temporal granularities with the data structure <$A_i$, $B_j(s)$, $v(s)$> $i \in {1, 2, \dots, n}$ and $j \in {1, 2, \dots, m}$. The framework involves faceting approach with one cyclic granularity plotted along the x-axis, the other one across facet and the dependent time series variable on the y-axis. -->

@@

Consider looking at letter value plots instead of quantile plots to look at the same data in \autoref{fig:bothcust-lv}. There is additional information that we can derive looking at the distribution.
There is bimodality in the early hours of the day, implying both low and high energy consumption is probable in the early hours of the day both for weekdays and weekends. Also hours from 7 to 13 look most volatile. If we visualize the same data with reverse mapping of the cyclic granularities, then the natural tendency would be to compare weekend and weekday behavior within each hour and not across hours. For example in \autoref{fig:cust-lv}, it can be seen that median energy consumption for the early morning hours is extremely high for weekdays compared to weekends. Also, outliers are more prominent in the latter part of the day. All of these indicates that looking at different distribution plots or changing the mapping might shed lights on different aspect of the energy behavior for the same customer.

@@

```{r granadvice, echo = TRUE}
smart_meter10 %>% 
  filter(customer_id %in% c(10017936))%>% 
  gran_advice("wknd_wday",
                            "hour_day")
```

@@

```{r cricex, fig.cap= "Letter value plot of runs per over across overs of the inning faceted by innings of the match. No upward shift in runs in the second innings like that in the first implying teams are more vulnerable to score more in the first innings as they approach the end of the inning."}

@@

```{r field, fig.cap = "Distribution of lagged run rate across overs of the innings and wickets dismissed due to fielding. This shows good fielding in previous over leads to lower run rate in the subsequent over. Also run rates are much less volatile in case there are at least one wicket in the previous over."}

>>>>

   Visualizing data across granularities which are either single-order-up or multiple-order-up or periodic/aperiodic helps us to understand periodicities, pattern and anomalies in the data. Because of the large volume of data available, using displays of probability distributions conditional on one or more granularities is a potentially useful approach. This work provides tools for creating granularities and exploring the associated time series within the tidy workflow, so that probability distributions can be examined using the range of graphics available in ggplot2 [@Wickham2009pk]. 
bibliography: [bibliography.bib]
preamble: >
  \usepackage{mathtools}
  \usepackage{amssymb} 
  \usepackage{booktabs}
  \usepackage{longtable}
output:
  bookdown::pdf_book:
    base_format: rticles::asa_article
    fig_height: 5
    fig_width: 8
    fig_caption: yes
    dev: "pdf"
---

@@

Calendar-based graphics [@wang2018calendar] are useful in visualizing patterns in the weekly and monthly structure well and are capable of checking the weekends or special days. Any sub-daily resolution temporal data can also be displayed using this type of faceting [@Wickham2009pk] with days of the week, month of the year and another sub-daily deconstruction of time. But calendar effects are not restricted to conventional day-of-week, month-of-year ways of deconstructing time. There can be several classes of time deconstructions, viz. based on the arrangement (linear vs. cyclic) or hierarchical order of the calendar. Linear time granularities respect the linear progression of time such as hours, days, weeks and months. One of the first attempts of characterizing these granularities occur in @Bettini1998-ed. The definitions and rules defined are inadequate to reflect periodicities in time. Hence, there is a need to define cyclic time granularities in a different approach, which can be useful in visualizing periodic behavior. Cyclic time granularities can be circular, quasi-circular or aperiodic. Examples of circular can be hour of the day, day of the week, that of quasi-circular can be day of the month and examples of aperiodic granularities can be public or school holidays. Time deconstructions can also be based on the hierarchical structure of time. For example, hours are nested within days, days within weeks, weeks within months, and so on. Hence, it is possible to construct single-order-up granularities like second of the minute or multiple-order-up granularities like second of the hour. lubridate [@G_Grolemund2011-vm] creates easy access and manipulation of common date-time objects. But most accessor functions are limited to single-order-up granularities.

@@

The grammar of graphics introduced a framework to construct statistical graphics by relating data space to the graphic space [@Wilkinson1999-nk]. The layered grammar of graphics proposed by [@Wickham2009pk], which is an alternate and modified parametrization of the grammar suggests that graphics are made up of distinct layers of grammatical elements. There are seven grammatical elements in total, out of which data, aesthetics and geometries are essential elements. Data are observations made on a number of variables, typically thought of as a sample from a population distribution. Aesthetic mapping describes how variables are mapped to elements of the plot, such as horizontal axis or color. Geometries determine the physical representation of the data in the plot. The remaining grammatical elements of a graphic namely facets, statistics, coordinates and themes are optional and contains detail of the plot. Facets dictates how to split up our plot and is a powerful tool for exploratory data analysis as we can rapidly compare patterns in different parts of the data and can see whether they are the same or different.

@@

<!-- We exploit some features of layers and faceting [@Wilkinson1999-nk] and [@Wickham2009pk] to come up with a recommendation system while visualizing the distribution of the "dependent" variable across bivariate cyclic temporal granularities with the data structure <$A_i$, $B_j(s)$, $v(s)$> $i \in {1, 2, \dots, n}$ and $j \in {1, 2, \dots, m}$. The framework involves faceting approach with one cyclic granularity plotted along the x-axis, the other one across facet and the dependent time series variable on the y-axis. -->

@@

Consider looking at letter value plots instead of quantile plots to look at the same data in \autoref{fig:bothcust-lv}. There is additional information that we can derive looking at the distribution.
There is bimodality in the early hours of the day, implying both low and high energy consumption is probable in the early hours of the day both for weekdays and weekends. Also hours from 7 to 13 look most volatile. If we visualize the same data with reverse mapping of the cyclic granularities, then the natural tendency would be to compare weekend and weekday behavior within each hour and not across hours. For example in \autoref{fig:bothcust-lv}, it can be seen that median energy consumption for the early morning hours is extremely high for weekdays compared to weekends. Also, outliers are more prominent in the latter part of the day. All of these indicates that looking at different distribution plots or changing the mapping might shed lights on different aspect of the energy behavior for the same customer.

@@

```{r granadvice, echo = TRUE, warning=FALSE, message=FALSE}
smart_meter10 %>% 
  filter(customer_id %in% c(10017936))%>% 
  gran_advice("wknd_wday",
                            "hour_day")
```

@@

```{r cricex, fig.cap= "Letter value plot of runs per over across overs of the inning faceted by innings of the match. No upward shift in runs in the second innings like that in the first implying teams are more vulnerable to score more in the first innings as they approach the end of the inning.", warning=FALSE, message = FALSE}

@@

```{r field, fig.cap = "Distribution of lagged run rate across overs of the innings and wickets dismissed due to fielding. This shows good fielding in previous over leads to lower run rate in the subsequent over. Also run rates are much less volatile in case there are at least one wicket in the previous over.", warning=FALSE}
